Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-09-19 ms.tex
more edits

The one-dimensional steady state flow of gas is usually characterized
by an inflow-outflow structure, with a critical radius known as the
``stagnation radius" $\rs$ where the radial velocity goes to zero.
Mass loss from stars interior to the stagnation radius is accreted,
while that outside $\rs$ is unbound in an outflow from the nucleus.
The stagnation radius, rather than the Bondi radius, is thus
ultimately responsible for setting the SMBH accretion rate.  Although
we find that $\rs$ typically lies within a factor of few of the
nominal Bondi radius, the difference in the accretion rate can be
significant.  Furthermore, in some cases we find that no stagnation
radius exists at all; this formally implies that the entire of ISM of
the galaxy is feeding the SMBH, potentially resulting in much higher
accretion rate than would be predicted by simple Bondi accretion.  The
main goal of our analysis is to provide a framework for interpreting
the properties of quiescent galactic nuclei given the relatively well
understood physics of stellar winds and supernova heating.

@@

Rewriting equation (\ref{eq:mdot_gas}) as
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{t_{\ff}} =\frac{4 \pi}{3} \rs^2 \rho(\rs) v_{\ff}(\rs)
\end{align}
we note the qualitative similarity between this expression and the Bondi formula (equation~[\ref{eq:bondi}]) with the exception that $\rb$ is replaced by $\rs$.  Using equation (\ref{eq:stag_simple})
\begin{align}
  \rs=\frac{7}{2}\frac{G \Mbh}{\vwO^2} \approx \frac{14}{3}\frac{GM_{\bullet}}{c_{s}^{2}}\frac{v_{w}^{2}}{v_{w,0}^{2}},
  \label{eq:rs_simple}
\end{align}
where in the second line we have used the fact that $v_{w}^{2}/2 =
c_s^2/(\gamma-1)$ at $\rs$.  Thus, provided that a stagnation radius
exists which is interior to the sphere of influence radius, the
accretion rate predicted by our model is quite close to $\dot{M}_{\rm
  B}$.
 %({\bf BDM: Check - as this depends on
 % whether eq 10 is correct})

@@

\subsubsection{Combined Heating Rate} The overall heating rate $\vwO$
has contributions from Type Ia SNes and main sequence winds.  The
relative contributions will depend on both the age of the stellar
population, $\tage$, and $\Mbh$.  This is illustrated in
Figure~\ref{fig:vweff}. For black hole masses below $10^7 \Msun$ SNe
Ia will not contribute to $\vwO$ as the flow time scale is short
compared to $t_{\rm Ia}$, and the heating is dominated by main
sequence winds (with $\vwO\simeq 100$ km/s).  On the other hand, for
$\Mbh \gsim 10^8 \Msun$, heating from Ias becomes important, as
$\rs>\rIa$. In this regime, we add the $\vwO$ values from SNe Ias and
from main sequence winds in quadrature. $\vwO$ from SNe Ia's depends
on the rate of mass loss ($\eta$), which is in turn a function of
$t_{\star}$. This dependence is why the effective $\vwO$ increases
with $\tage$ for large $\Mbh$.

@@

  \begin{figure}
    \includegraphics[width=\columnwidth]{vw-contour.eps}
    \caption{\label{fig:vweff} Combined heating rate $\vwO$ of the CNM
      in the parameter space of SMBH mass $\Mbh$ and age of the
      stellar population $t_{\star}$.  We assume that stellar winds
      contribute a constant heating rate $100$ km/s. The contribution
      from SNe Ia is calculated using equation (\ref{eq:vw_sne}). The
      contribution from SNe Ia is only added if the $r_s$
      corresponding to the resulting $\vwO$ is outside
      $\rIa$.}%describe other cases here.
  \end{figure}

@@

This is shown in the top panel Figure~\ref{fig:bh_xray}. The
relationship from \citealt{MillerGallo+:2014a} and their detection
upper limits are the black and blue solid lines
respectively. The points correspond to the $\dot{M} c^2$ for our
solutions (scaled by $5\times 10^{-7}$).  We must choose a $\vwO$ for
each galaxy.  We select $\vwO$=200 km/s if $\rs<\rIa$ for this $\vwO$,
$\vwO$=500 km/s if $\rs>\rIa$ for this $\vwO$, or exclude it. If
$L_x \sim \Mdot^2$ the discrepancy becomes worse, as shown in the
bottom panel of Fig.~\ref{fig:bh_xray}, where we plot $2 \times
10^{-4} (\eddr) \Mdot c^2$ from our solutions with the relation from
\citealt{MillerGallo+:2014a}.

@@

Note that we have two free parameters ($\eta$ and $\vwO$) in our
model, which implies that we can always match the observed temperature
and densities at at least one point.  As a general rule, we find that
a wind heating parameter $\vwO=500$ km s$^{-1}$ (close to, but
somewhat higher than that expected from SNe Ia) gives rough agreement
with the observed temperatures.  We now highlight results for specific
galaxies.
% Thus, we choose $\vwO=500$ km/s.  We choose an
% $\eta$ so that our solution would be consistent with the observed
% density profiles. The comparison is shown in
%% AG-The chosen properties, in particular eta may be summarized in a table.
%% AG-Mention rising temperature profiles on large scales.

@@

%It would be nice to have error bars for the observational points in this plot...
\begin{figure*}
  \includegraphics[width=\columnwidth]{NGC5813_T.eps}
  \includegraphics[width=\columnwidth]{NGC5813_rho.eps}
  \caption{\label{fig:allen_compare} Comparison of the radial profiles of electron number density $n_e$ and temperature $T$ to
    the measured radial profiles provided in \citet{AllenDunn+:2006a} and \citet{ RussellMcNamara+:2013a}}.
  %% AG in the future we will have to be carfeful of the mean molecular weight.
\end{figure*}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{bh_xray.eps}
\caption{\label{fig:bh_xray} X-ray luminosity plotted as a function of
  stellar mass for our solutions (points), the trend of nuclear
  x-ray luminosity with stellar mass (black line) and the detection
  upper limit (blue line) from Fig. 1 of
  \citealt{MillerGallo+:2014a}. In the top panel we take the X-ray
  luminosity to be $\epsilon_1 \Mdot c^2$, with $\epsilon_1=5\times
  10^{-7}$ chosen to make our luminosities comparable to those
  observed. In the bottom panel, we take the
  accretion power to be quadratic in $\Mdot$. $L_x=\epsilon_2 \eddr \Mdot
  c^2$ with $\epsilon2$=$2\times10^{-4}$. For a given galaxy, we take
  either $\vwO=200$ km/s (green points) or $\vwO=500$ km/s (orange
  points). We choose the solution for which the relative positions of
  $\rs$ and $\rIa$ are consistent with the assumed level of heating}
\end{figure}

@@

  \begin{enumerate}
  \item We find that $\eddr \propto \Mbh^{\alpha}$, where the value of
    $\alpha$ depends on the slope of the stellar density profile
    ($\alpha\simeq0.5,1$ for cusp and core galaxies respectively). One
    possible solution that the wind mass loss rate (i.e. $\eta$)
    decreases with $\Mbh$, as would be expected if low mass SMBH
    preferentially hosted older stellar populations. Another possible
    explanation is that the effective heating rate of the CNM $v_{\rm w,0}$
    decreases with decreasing SMBH mass. This could again be due to an
    older stellar population, since $v_{\rm w,0} \propto \eta^{−1/2}$
    for heating sources such as SN Ia and millisecond pulsars.
  \item We compare the $\Mdot$ for each of our solution to the one
    calculated from the Bondi formula. We find that they generally to
    a factor of a few. In our model the accretion rate is determined
    from the gas properties at the stagnation radius, $\rs$, which
    agrees with the Bondi radius, $\rb$, within a factor of a
    few. Thus, use of the Bondi formula is a good approximation.
  \item We compare our solutions for density and temperature to those
    inferred from Chandra x-ray observations by \citealt{AllenDunn+:2006a}
    and \citealt{RussellMcNamara+:2013a} for a handful of galaxies in
    which our sample overlap with theirs. In general, we cannot match
    the observed gas density slope on large scales. 
  \end{enumerate}

>>>>

The one-dimensional steady state flow of gas is usually characterized
by an inflow-outflow structure, with a critical radius known as the
``stagnation radius" $\rs$ where the radial velocity goes to zero.
Mass loss from stars interior to the stagnation radius is accreted,
while that outside $\rs$ is unbound in an outflow from the nucleus.
The stagnation radius, rather than the Bondi radius, is thus
ultimately responsible for setting the SMBH accretion rate, although
we find that $\rs$ typically lies within a factor of few of the
nominal Bondi radius.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion. 
The main goal of our analysis is to provide a framework for
interpreting the properties of quiescent galactic nuclei given the
relatively well understood physics of stellar winds and supernova
heating.

@@

Rewriting equation (\ref{eq:mdot_gas}) as
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{t_{\ff}} =\frac{4 \pi}{3} \rs^2 \rho(\rs) v_{\ff}(\rs)
\end{align}
we note the qualitative similarity between this expression and the Bondi formula (equation~[\ref{eq:bondi}]) with the exception that $\rb$ is replaced by $\rs$.  Using equation (\ref{eq:stag_simple})
\begin{align}
  \rs\approx\frac{7}{2}\frac{G \Mbh}{\vwO^2} \approx \frac{14}{3}\frac{GM_{\bullet}}{c_{s}^{2}},
  \label{eq:rs_simple}
\end{align}
where in the second line we have used the fact that $v_{w}^{2}/2 =
c_s^2/(\gamma-1)$ at $\rs$.  Thus, provided that a stagnation radius
exists which is interior to the sphere of influence radius, the
accretion rate predicted by our model is quite close to $\dot{M}_{\rm
  B}$.
 %({\bf BDM: Check - as this depends on
 % whether eq 10 is correct})

@@

\subsubsection{Combined Heating Rate} The overall heating rate $\vwO$
has contributions from Type Ia SNes and main sequence winds.  The
relative contributions will depend on both the age of the stellar
population, $\tage$, and $\Mbh$.  This is illustrated in
Figure~\ref{fig:vweff}. For black hole masses below $10^7 \Msun$ SNe
Ia will not contribute to $\vwO$ as the flow time scale is short
compared to $t_{\rm Ia}$, and the heating is dominated by main
sequence winds (with $\vwO\simeq 100-150$ km/s), and possibly
millisecond pulsars ($\vwO\simeq 120\epsilon^{1/2} (\eta/0.1)^{-1/2}$
km s$^{-1}$).  Optimistically, we would obtain $\vwO\simeq 200$ km/s
from the combined effects of MSPs and main sequence winds. On the
other hand, for $\Mbh \gsim 10^8 \Msun$, heating from Ias becomes
important, as $\rs>\rIa$. In this regime, we add the $\vwO$ values
from SNe Ias and from main sequence winds in quadrature. $\vwO$ from
SNe Ia's depends on the rate of mass loss ($\eta$), which is in turn a
function of $t_{\star}$. This dependence is why the effective $\vwO$
increases with $\tage$ for large $\Mbh$.

@@

  \begin{figure}
    \includegraphics[width=\columnwidth]{vw-contour.eps}
    \caption{\label{fig:vweff} Combined heating rate $\vwO$ of the CNM
      in the parameter space of SMBH mass $\Mbh$ and age of the
      stellar population $t_{\star}$.  We assume that stellar winds
      contribute a constant heating rate $100$ km/s. The contribution
      from SNe Ia is calculated using equation (\ref{eq:vw_sne}), with
      thermalization efficiency $\epsilon_{\rm Ia}=0.4$. The
      contribution from SNe Ia is only added if the $r_s$
      corresponding to the resulting $\vwO$ is outside
      $\rIa$.}%describe other cases here.
  \end{figure}

@@

This is shown in the top panel Figure~\ref{fig:bh_xray}. The
relationship from \citealt{MillerGallo+:2014a} and their detection
upper limits are the black and blue solid lines
respectively. The points correspond to the $\dot{M} c^2$ for our
solutions (scaled by $5\times 10^{-7}$).  We must choose a $\vwO$ for
each galaxy.  We select $\vwO$=200 km/s, roughly the expected level of
heating expected from main sequence the stellar winds and MSPs. If
$L_x \sim \Mdot^2$ the discrepancy becomes worse, as shown in the
bottom panel of Fig.~\ref{fig:bh_xray}, where we plot $2 \times
10^{-4} (\eddr) \Mdot c^2$ from our solutions with the relation from
\citealt{MillerGallo+:2014a}.

@@

Note that we have two free parameters ($\eta$ and $\vwO$) in our
model, which implies that we can always match the observed temperature
and densities at at least one point.  As a general rule, we find that
a wind heating parameter $\vwO=500$ km s$^{-1}$ gives rough agreement
with the observed temperatures.  We now highlight results for specific
galaxies.
% Thus, we choose $\vwO=500$ km/s.  We choose an
% $\eta$ so that our solution would be consistent with the observed
% density profiles. The comparison is shown in
%% AG-The chosen properties, in particular eta may be summarized in a table.
%% AG-Mention rising temperature profiles on large scales.

@@

%It would be nice to have error bars for the observational points in this plot...
\begin{figure*}
  \includegraphics[width=\columnwidth]{NGC5813_T.eps}
  \includegraphics[width=\columnwidth]{NGC5813_rho.eps}
  \caption{\label{fig:allen_compare} Comparison of the radial profiles
    (blue lines) of electron number density $n_e$ and temperature $T$
    to the measured values (black square) in \citet{AllenDunn+:2006a}
    and \citet{ RussellMcNamara+:2013a}. The dashed blue lines are
    power law extrapolations of our profiles. Note that in going from
    $\rho$ to $n_e$ we our adopted value for the molecular weight $\mu=1$.}
  %% AG in the future we will have to be carfeful of the mean molecular weight.
\end{figure*}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{bh_xray.eps}
\caption{\label{fig:bh_xray} X-ray luminosity plotted as a function of
  stellar mass for our solutions (points), the trend of nuclear x-ray
  luminosity with stellar mass (black line) and the detection upper
  limit (blue line) from Fig. 1 of \citealt{MillerGallo+:2014a}. In
  the top panel we take the X-ray luminosity to be $\epsilon_1 \Mdot
  c^2$, with $\epsilon_1=5\times 10^{-7}$ chosen to make our
  luminosities comparable to those observed. In the bottom panel, we
  take the accretion power to be quadratic in $\Mdot$. $L_x=\epsilon_2
  \eddr \Mdot c^2$ with $\epsilon2$=$2\times10^{-4}$. For each galaxy
  we choose the $\vwO=200$ km/s solution.}
\end{figure}

@@

  \begin{enumerate}
  \item We find that $\eddr \propto \Mbh^{\alpha}$, where the value of
    $\alpha$ depends on the slope of the stellar density profile
    ($\alpha\simeq0.5,1$ for cusp and core galaxies respectively). One
    possible solution that the wind mass loss rate (i.e. $\eta$)
    decreases with $\Mbh$, as would be expected if low mass SMBH
    preferentially hosted older stellar populations. Another possible
    explanation is that the effective heating rate of the CNM $v_{\rm w,0}$
    decreases with decreasing SMBH mass. This could again be due to an
    older stellar population, since $v_{\rm w,0} \propto \eta^{−1/2}$
    for heating sources such as SN Ia and millisecond pulsars.
  \item We compare the $\Mdot$ for each of our solution to the one
    calculated from the Bondi formula. We find that they generally to
    a factor of a few. In our model the accretion rate is determined
    from the gas properties at the stagnation radius, $\rs$, which
    agrees with the Bondi radius, $\rb$, within a factor of a
    few. Thus, use of the Bondi formula is a good approximation.
  \item We compare our solutions for density and temperature to those
    inferred from Chandra x-ray observations by \citet{AllenDunn+:2006a}
    and \citet{RussellMcNamara+:2013a} for a handful of galaxies in
    which our sample overlap with theirs. In general, we cannot match
    the observed gas density slope on large scales. 
  \end{enumerate}
